---
title: "Adoption Metadata Analysis"
author: "Chiranjit Mukherjee"
date: "12/14/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(stargazer)
```

<br>

Setting Seed
```{r}
set.seed(12345)
```



### Input metadata files
```{r}
ISR_sal_meta_mch_mo <- read.table(file="ISR_sal_meta_mch_mo.txt", header = T, sep="\t")
ISR_sal_meta_mch_ch <- read.table(file="ISR_sal_meta_mch_ch.txt", header = T, sep="\t")
```



## ISR Sal M-Ch Multiple Linear Regression Analysis for unbalanced metadata
<br.
```{r}
# Check dimensions
dim(ISR_sal_meta_mch_mo) # 103 x 24
dim(ISR_sal_meta_mch_ch) # 103 x 24 (distances added)

# Create new field Caries_yn (yes/no)
ISR_sal_meta_mch_mo$caries_yn <- "No"
ISR_sal_meta_mch_mo$caries_yn[ISR_sal_meta_mch_mo$caries_level > 0] <- "Yes"

# Mother's caries_yn
ISR_sal_meta_mch_ch$mo_caries <- ISR_sal_meta_mch_mo$caries_yn[match(ISR_sal_meta_mch_ch$family_id, ISR_sal_meta_mch_mo$family_id)]

# Mother's gingivitis
ISR_sal_meta_mch_ch$mo_ging <- ISR_sal_meta_mch_mo$gingivitis[match(ISR_sal_meta_mch_ch$family_id, ISR_sal_meta_mch_mo$family_id)]

# Mother's plaque level
ISR_sal_meta_mch_ch$mo_plq_level <- ISR_sal_meta_mch_mo$plaque_level[match(ISR_sal_meta_mch_ch$family_id, ISR_sal_meta_mch_mo$family_id)]

# Mother's ethnicity
ISR_sal_meta_mch_ch$mo_eth <- ISR_sal_meta_mch_mo$ethnicity[match(ISR_sal_meta_mch_ch$family_id, ISR_sal_meta_mch_mo$family_id)]

# Mother's tongue biofilm
ISR_sal_meta_mch_ch$mo_tng_bio <- ISR_sal_meta_mch_mo$tongue_biofilm[match(ISR_sal_meta_mch_ch$family_id, ISR_sal_meta_mch_mo$family_id)]

# Mother's antibiotics
ISR_sal_meta_mch_ch$mo_anti <- ISR_sal_meta_mch_mo$antibiotics_past_month[match(ISR_sal_meta_mch_ch$family_id, ISR_sal_meta_mch_mo$family_id)]

# Mother's race
ISR_sal_meta_mch_ch$mo_race <- ISR_sal_meta_mch_mo$race[match(ISR_sal_meta_mch_ch$family_id, ISR_sal_meta_mch_mo$family_id)]

# ethnicity data not available for most children
ISR_sal_meta_mch_ch$ethnicity <- NULL

```

```{r}
# Age

# Stats
set.seed(12345);wilcox.test(ISR_sal_meta_mch_ch$age ~ ISR_sal_meta_mch_ch$status)$p.value # p-value = 0.2937314 (Balanced)

# Gender
# Check levels
ISR_sal_meta_mch_ch$gender %>% unique()

# Check table
ISR_sal_meta_mch_ch$gender  %>% table()

# Contingency Table
gender_table <- table(ISR_sal_meta_mch_ch$gender, ISR_sal_meta_mch_ch$status)

# Check C table
gender_table
#           Adopted Biological
# Female      17         29
# Male        32         25

# Fisher test
set.seed(12345);round(fisher.test(gender_table)$p.value,3) # p-value = 0.074 (Balanced)


# Delivery Mode
ISR_sal_meta_mch_ch$del_mode %>% unique()

# Fix NA
ISR_sal_meta_mch_ch$del_mode <- gsub("NA", NA, ISR_sal_meta_mch_ch$del_mode)
ISR_sal_meta_mch_ch$del_mode <- gsub("Unknown", NA, ISR_sal_meta_mch_ch$del_mode)

# Check table
ISR_sal_meta_mch_ch$del_mode  %>% table()

# Contingency Table
del_mode_table <- table(ISR_sal_meta_mch_ch$del_mode, ISR_sal_meta_mch_ch$status)

# Check C table
del_mode_table
#           Adopted Biological
# C-Section      12         19
# Vaginally      33         35

# Fisher test
set.seed(12345);round(fisher.test(del_mode_table)$p.value, 3) # p-value = 0.392 (Balanced)


# Feeding Mode
ISR_sal_meta_mch_ch$feed_mode %>% unique()

# Fix NA
ISR_sal_meta_mch_ch$feed_mode <- gsub("NA", NA, ISR_sal_meta_mch_ch$feed_mode)
ISR_sal_meta_mch_ch$feed_mode <- gsub("Unknown", NA, ISR_sal_meta_mch_ch$feed_mode)

# Check table
ISR_sal_meta_mch_ch$feed_mode  %>% table()

# Contingency Table
feed_mode_table <- table(ISR_sal_meta_mch_ch$feed_mode, ISR_sal_meta_mch_ch$status)

# Check table
feed_mode_table
#               Adopted Biological
# Breast-fed        0         21
# Combination       1         27
# Formula-fed      46          6

# Fisher test
set.seed(12345);round(fisher.test(feed_mode_table)$p.value, 3) # 0 (Unbalanced)



# Race
ISR_sal_meta_mch_ch$race %>% unique()
# "White"                        "White and Asian"              "White and Arabic"             "Black or African American"    "Other (both black and white)"

# Fix names
ISR_sal_meta_mch_ch$race <- gsub("White and Asian", "Mixed", ISR_sal_meta_mch_ch$race)
ISR_sal_meta_mch_ch$race <- gsub("White and Arabic", "Mixed", ISR_sal_meta_mch_ch$race)
ISR_sal_meta_mch_ch$race <- gsub("Other \\(both black and white\\)", "Mixed", ISR_sal_meta_mch_ch$race)

# Check table
ISR_sal_meta_mch_ch$race  %>% table()

# Contingency Table
race_table <- table(ISR_sal_meta_mch_ch$race, ISR_sal_meta_mch_ch$status)

# Check table
race_table
#                             Adopted   Biological
# Black or African American      17          1
# Mixed                           8          5
# White                          24         48

# Fisher test
set.seed(12345);round(fisher.test(race_table)$p.value, 3) # p-value = 0 (Unbalanced)




# Plaque Level
ISR_sal_meta_mch_ch$plaque_level %>% unique()
# "N/A"      "Moderate" "Mild"     "None"     "NA"

# Fix NA
ISR_sal_meta_mch_ch$plaque_level <- gsub("N/A", NA, ISR_sal_meta_mch_ch$plaque_level)
ISR_sal_meta_mch_ch$plaque_level <- gsub("NA", NA, ISR_sal_meta_mch_ch$plaque_level)


# Check table
ISR_sal_meta_mch_ch$plaque_level  %>% table()

# Contingency Table
plaque_level_table <- table(ISR_sal_meta_mch_ch$plaque_level, ISR_sal_meta_mch_ch$status)

# Check table
plaque_level_table
#            Adopted Biological
# Mild          17         27
# Moderate       3          4
# None          25         21


# Fisher test
set.seed(12345);round(fisher.test(plaque_level_table)$p.value, 3) # p-value = 0.306 (Balanced)



# Tongue Biofilm
ISR_sal_meta_mch_ch$tongue_biofilm %>% unique()
# "NA"    "Light" "None"  "Heavy"

# Fix NA
ISR_sal_meta_mch_ch$tongue_biofilm <- gsub("NA", NA, ISR_sal_meta_mch_ch$tongue_biofilm)

# Check table
ISR_sal_meta_mch_ch$tongue_biofilm  %>% table()

# Contingency Table
tongue_biofilm_table <- table(ISR_sal_meta_mch_ch$tongue_biofilm, ISR_sal_meta_mch_ch$status)

# Check table
tongue_biofilm_table
#            Adopted Biological
# Heavy       1          0
# Light      21          6
# None       24         46


# Fisher test
set.seed(12345);fisher.test(tongue_biofilm_table)$p.value # p-value = 0.0001348483 (Unbalanced)



# Gingivitis
ISR_sal_meta_mch_ch$gingivitis %>% unique()
# "NA"       "Moderate" "None"     "Mild"

# Fix NA
ISR_sal_meta_mch_ch$gingivitis <- gsub("NA", NA, ISR_sal_meta_mch_ch$gingivitis)


# Check table
ISR_sal_meta_mch_ch$gingivitis %>% table()

# Contingency Table
gingivitis_table <- table(ISR_sal_meta_mch_ch$gingivitis, ISR_sal_meta_mch_ch$status)

# Check table
gingivitis_table
#            Adopted Biological
# Mild          11         25
# Moderate       2          2
# None          33         25


# Fisher test
set.seed(12345);round(fisher.test(gingivitis_table)$p.value, 3) # p-value = 0.0326 (Unbalanced)



# Caries Level
range(ISR_sal_meta_mch_ch$caries_level)
# 0 11

# Create new field Caries_yn (yes/no)
ISR_sal_meta_mch_ch$caries_yn <- "No"
ISR_sal_meta_mch_ch$caries_yn[ISR_sal_meta_mch_ch$caries_level > 0] <- "Yes"

# Check table
ISR_sal_meta_mch_ch$caries_yn %>% table()

# Contingency Table
caries_yn_table <- table(ISR_sal_meta_mch_ch$caries_yn, ISR_sal_meta_mch_ch$status)

# Check table
caries_yn_table
#        Adopted  Biological
# No       42         46
# Yes       7          8

# Fisher test
set.seed(12345);round(fisher.test(caries_yn_table)$p.value, 3) # p-value = 1 (Balanced)


# Daycare

# School: Yes or No
ISR_sal_meta_mch_ch$School <- as.character(ISR_sal_meta_mch_ch$School)
ISR_sal_meta_mch_ch$School %>% unique()
# "Daycare"      "None"         "NA"           "Pre School"   "kindergarten" "School"   

# Replace NA with MD for missing data
ISR_sal_meta_mch_ch$School[is.na(ISR_sal_meta_mch_ch$School)] <- "MD" # Missing data


# Making new field for Daycare and setting as yes/no field
for (i in 1:nrow(ISR_sal_meta_mch_ch)){
  if (ISR_sal_meta_mch_ch$School[i] == "MD"){
    ISR_sal_meta_mch_ch$daycare[i] = NA
  }
  else if (ISR_sal_meta_mch_ch$School[i] == "None"){
    ISR_sal_meta_mch_ch$daycare[i] = "No"
  }
  else {
    ISR_sal_meta_mch_ch$daycare[i] <- "Yes"
  }
}

# Daycare levels
ISR_sal_meta_mch_ch$daycare  %>% table()

# Contingency table
daycare_table <- table(ISR_sal_meta_mch_ch$daycare, ISR_sal_meta_mch_ch$status)

# Check C table
daycare_table
#       Adopted   Biological
# No       20         15
# Yes      29         36

# Fisher test
set.seed(12345);round(fisher.test(daycare_table)$p.value, 3) # p-value = 0.295 (Balanced)




# antibiotics_past_month
ISR_sal_meta_mch_ch$antibiotics_past_month %>% unique()
# "No"  "Yes"

# Check table
ISR_sal_meta_mch_ch$antibiotics_past_month %>% table()

# Contingency Table
antibiotics_past_month_table <- table(ISR_sal_meta_mch_ch$antibiotics_past_month, ISR_sal_meta_mch_ch$status)

# Check table
antibiotics_past_month_table
#        Adopted Biological
# No       47         49
# Yes       2          5

# Fisher test
set.seed(12345);round(fisher.test(antibiotics_past_month_table)$p.value, 3) # p-value = 0.441 (Balanced)



# Mother's age
# Add mother's age for children
ISR_sal_meta_mch_ch$mo_age <- ISR_sal_meta_mch_mo$age[match(ISR_sal_meta_mch_ch$family_id, ISR_sal_meta_mch_mo$family_id)]

boxplot(ISR_sal_meta_mch_ch$mo_age ~ ISR_sal_meta_mch_ch$status)

# Stats
set.seed(12345); wilcox.test(ISR_sal_meta_mch_ch$mo_age ~ ISR_sal_meta_mch_ch$status)$p.value # p-value = 2.122e-06 (Unbalanced)




# Mother's gingivitis
ISR_sal_meta_mch_ch$mo_ging %>% unique()
# "NA"       "Moderate" "None"     "Mild"

# Fix NA
ISR_sal_meta_mch_ch$mo_ging <- gsub("NA", NA, ISR_sal_meta_mch_ch$mo_ging)

# Check table
ISR_sal_meta_mch_ch$mo_ging %>% table()

# Contingency Table
mo_ging_table <- table(ISR_sal_meta_mch_ch$mo_ging, ISR_sal_meta_mch_ch$status)

# Check table
mo_ging_table
#            Adopted Biological
# Mild          25         14
# Moderate       7          5
# None          16         34

# Fisher test
set.seed(12345);round(fisher.test(mo_ging_table)$p.value, 3) # p-value = 0.007327 (Balanced)


# mo_plq_level
ISR_sal_meta_mch_ch$mo_plq_level %>% table()
# Mild Moderate      N/A     none     None 
#  57        7        2        1       37 

# Fix NA
ISR_sal_meta_mch_ch$mo_plq_level <- gsub("N/A", NA, ISR_sal_meta_mch_ch$mo_plq_level)
ISR_sal_meta_mch_ch$mo_plq_level <- gsub("none", "None", ISR_sal_meta_mch_ch$mo_plq_level)

# Contingency Table
mo_plq_level_table <- table(ISR_sal_meta_mch_ch$mo_plq_level, ISR_sal_meta_mch_ch$status)

# Check table
mo_plq_level_table
#            Adopted   Biological
# Mild          34         23
# Moderate       5          2
# None           9         28

# Fisher test
set.seed(12345);round(fisher.test(mo_plq_level_table)$p.value, 3) # p-value = 0.001 (Unbalanced)



# mo_eth
ISR_sal_meta_mch_ch$mo_eth %>% table()
# Hispanic     Hispanic or Latino Non Hispanic or Latino 
# 1                      1                    102 

# Fix NA
ISR_sal_meta_mch_ch$mo_eth <- gsub("Hispanic", "Hispanic or Latino", ISR_sal_meta_mch_ch$mo_eth)
ISR_sal_meta_mch_ch$mo_eth <- gsub("or Latino or Latino", "or Latino", ISR_sal_meta_mch_ch$mo_eth)

# Contingency Table
mo_eth_table <- table(ISR_sal_meta_mch_ch$mo_eth, ISR_sal_meta_mch_ch$status)

# Check table
mo_eth_table
#                          Adopted   Biological
# Hispanic or Latino           0          1
# Non Hispanic or Latino      49         52

# Fisher test
set.seed(12345);round(fisher.test(mo_eth_table)$p.value, 3) # p-value = 0.4963 (Balanced)



# mo_tng_bio
ISR_sal_meta_mch_ch$mo_tng_bio %>% table()
#    Heavy Light    NA  None 
#      2    50     2    50

# Fix NA
ISR_sal_meta_mch_ch$mo_tng_bio <- gsub("NA", NA, ISR_sal_meta_mch_ch$mo_tng_bio)

# Contingency Table
mo_tng_bio_table <- table(ISR_sal_meta_mch_ch$mo_tng_bio, ISR_sal_meta_mch_ch$status)

# Check table
mo_tng_bio_table
#         Adopted   Biological
# Heavy       1          1
# Light      32         18
# None       15         34

# Fisher test
set.seed(12345);round(fisher.test(mo_tng_bio_table)$p.value, 3) # p-value = 0.002 (Unbalanced)




# mo_anti
ISR_sal_meta_mch_ch$mo_anti %>% table()
# NA  No Yes 
#  1  92  10 

# Fix NA
ISR_sal_meta_mch_ch$mo_anti <- gsub("NA", NA, ISR_sal_meta_mch_ch$mo_anti)

# Contingency Table
mo_anti_table <- table(ISR_sal_meta_mch_ch$mo_anti, ISR_sal_meta_mch_ch$status)

# Check table
mo_anti_table
#         Adopted   Biological
# No       47         45
# Yes       2          8

# Fisher test
set.seed(12345);round(fisher.test(mo_anti_table)$p.value, 3) # p-value = 0.09472 (Balanced)


# mo_race
ISR_sal_meta_mch_ch$mo_race %>% unique()
# "White"                        "Asian"              "Black or African American"

# Check table
ISR_sal_meta_mch_ch$mo_race  %>% table()

# Contingency Table
mo_race_table <- table(ISR_sal_meta_mch_ch$mo_race, ISR_sal_meta_mch_ch$status)

# Check table
mo_race_table
#                             Adopted   Biological
# Asian                           0          3
# Black or African American       1          1
# White                          48         50

# Fisher test
set.seed(12345);round(fisher.test(mo_race_table)$p.value, 3) # p-value = 0.2444 (Balanced)



# Race match

ISR_sal_meta_mch_ch$race_match <- "Yes"
ISR_sal_meta_mch_ch$race_match[ISR_sal_meta_mch_ch$race != ISR_sal_meta_mch_ch$mo_race] <- "No"

# Contingency Table
race_match_table <- table(ISR_sal_meta_mch_ch$race_match, ISR_sal_meta_mch_ch$status)

# Check table
race_match_table
#       Adopted   Biological
# No       24          5
# Yes      25         49

# Fisher test
set.seed(12345);fisher.test(race_match_table)$p.value # p-value = 7.631e-06 (Unblanced)


# The unbalanced variables are: feed_mode, race, tongue_biofilm, gingivitis, mo_age, mo_plq_level, mo_tng_bio, race_match

# Subset
ISR_sal_meta_mch_ch_unb <- ISR_sal_meta_mch_ch[,c("sample", "status", "feed_mode", "race", "tongue_biofilm", "gingivitis", "mo_age","mo_ging", "mo_plq_level", "mo_tng_bio", "race_match", "dist")]
#View(ISR_sal_meta_mch_ch_unb)
dim(ISR_sal_meta_mch_ch_unb) # 103 x 11

# Output
#write.table(ISR_sal_meta_mch_ch_unb, file="Intermed_files/ISR_sal_meta_mch_ch_unb.txt", sep = "\t", quote = F, row.names = FALSE)
#write.table(ISR_sal_meta_mch_ch, file="Intermed_files/ISR_sal_meta_mch_ch_updated.txt", sep = "\t", quote = F, row.names = FALSE)
#write.table(ISR_sal_meta_mch_mo, file="Intermed_files/ISR_sal_meta_mch_mo_updated.txt", sep = "\t", quote = F, row.names = FALSE)
```




## Metadata Confounder Analysis

#### Refer to ISR Sal Metadata Analysis.Rmd for identificationo of unbalanced metadata

<br>
<br>

### Model Building with Unbalanced Clinical/Demographic Variables

<br>

#### Multiple Linear Regression Based Counfounder Analysis
<b>
```{r}

# Select only complete cases (exclude NAs) for multiple regression
ISR_sal_meta_mch_ch_unb <- na.omit(ISR_sal_meta_mch_ch_unb)
dim(ISR_sal_meta_mch_ch_unb) # 93  x 11
#View(ISR_sal_meta_mch_ch_unb)


# Remove sample column
ISR_sal_meta_mch_ch_unb$sample <- NULL

# Convert to factor and then to numeric for regression analysis
ISR_sal_meta_mch_ch_unb$status <- as.numeric(as.factor(ISR_sal_meta_mch_ch_unb$status)) # Levels: Adopted Biological
ISR_sal_meta_mch_ch_unb$feed_mode <- as.numeric(as.factor(ISR_sal_meta_mch_ch_unb$feed_mode)) # Levels: Breast-fed Combination Formula-fed
ISR_sal_meta_mch_ch_unb$race <- as.numeric(as.factor(ISR_sal_meta_mch_ch_unb$race)) # Levels: Black or African American Mixed White
ISR_sal_meta_mch_ch_unb$gingivitis <- as.numeric(factor(ISR_sal_meta_mch_ch_unb$gingivitis, levels =  c("None", "Mild", "Moderate"))) 
ISR_sal_meta_mch_ch_unb$tongue_biofilm <- as.numeric(factor(ISR_sal_meta_mch_ch_unb$tongue_biofilm, levels =  c("None", "Light", "Heavy")))
ISR_sal_meta_mch_ch_unb$mo_tng_bio <- as.numeric(factor(ISR_sal_meta_mch_ch_unb$mo_tng_bio, levels =  c("None", "Light", "Heavy")))
ISR_sal_meta_mch_ch_unb$mo_plq_level <- as.numeric(factor(ISR_sal_meta_mch_ch_unb$mo_plq_level, levels =  c("None", "Mild", "Moderate")))
ISR_sal_meta_mch_ch_unb$race_match <- as.numeric(as.factor(ISR_sal_meta_mch_ch_unb$race_match)) # Levels: No Yes

# Multiple linear regression analysis
(ISR_sal_meta_mch_ch_unb.fit <- lm(dist ~ status + feed_mode + race + mo_age + gingivitis + tongue_biofilm + mo_tng_bio + mo_plq_level + race_match, data=ISR_sal_meta_mch_ch_unb))
(ISR_sal_meta_mch_ch_unb.fit.sum <- summary(ISR_sal_meta_mch_ch_unb.fit)) # none significant

# Making table for output
ISR_sal_meta_mch_ch_unb.fit_tbl <- data.frame(ISR_sal_meta_mch_ch_unb.fit.sum$coefficients)
colnames(ISR_sal_meta_mch_ch_unb.fit_tbl) <- c("Estimates", "Std. Error", "t-value", "p-value")
#savView(ISR_sal_meta_mch_ch_unb.fit_tbl)
#write.table(ISR_sal_meta_mch_ch_unb.fit_tbl, file="/output/LM_table_sal_ISR_final.txt", sep="\t", row.names = T)

```

# New confounding checking
```{r,results="asis"}
# Multiple linear regression analysis
fit0<-lm(dist ~ status, data=ISR_sal_meta_mch_ch_unb)
fit1<-lm(dist ~ status+feed_mode, data=ISR_sal_meta_mch_ch_unb)
fit2<-lm(dist ~ status+race, data=ISR_sal_meta_mch_ch_unb)
fit3<-lm(dist ~ status+mo_age, data=ISR_sal_meta_mch_ch_unb)
fit4<-lm(dist ~ status+gingivitis, data=ISR_sal_meta_mch_ch_unb)
fit5<-lm(dist ~ status+tongue_biofilm, data=ISR_sal_meta_mch_ch_unb)
fit6<-lm(dist ~ status+mo_ging, data=ISR_sal_meta_mch_ch_unb)
fit7<-lm(dist ~ status+mo_tng_bio, data=ISR_sal_meta_mch_ch_unb)
fit8<-lm(dist ~ status+mo_plq_level, data=ISR_sal_meta_mch_ch_unb)
fit9<-lm(dist ~ status+race_match, data=ISR_sal_meta_mch_ch_unb)
fit10<-lm(dist ~ status + feed_mode + race + mo_age + gingivitis + tongue_biofilm +mo_ging+mo_tng_bio + mo_plq_level + race_match, data=ISR_sal_meta_mch_ch_unb)

stargazer(header=F,fit0,fit1,fit2,fit3,fit4,fit5,fit6,fit7,fit8,fit9,fit10,type="html",report="vcp*",title="Check confounding")


```